---
title: "2_reproduce_paper_code"
editor_options: 
  chunk_output_type: console
---

# Load libraries
```{r}
library(dplyr)
library(tidyr)
library(readr)
library(GenomicRanges)
library(reshape2)
library(openxlsx)
library(ggplot2)
library(Matrix)
library(EpiTrace)
library(Seurat)
library(SeuratObject)
library(ggtree)
library(EnsDb.Hsapiens.v86)
library(patchwork)
library(ArchR)
library(Signac)
library(ggbeeswarm)
library(ggthemes)
library(ggsci)
library(ggpubr)
library(pheatmap)
library(TxDb.Hsapiens.UCSC.hg19.knownGene)
library(ChIPseeker)
library(ComplexHeatmap)
library(zoo)
library(RColorBrewer)
library(openxlsx)
library(WGCNA)
library(parallel)
library(TxDb.Hsapiens.UCSC.hg38.knownGene)
library(ComplexHeatmap)
library(circlize)

options(bitmapType='cairo') # solve the PNG/X11 issue
data("clock_gr_list")
```


# Running paper sample code
```{r}
path = '/projectnb/ds596/projects/Team 3/paper_code'

# build Epitrace object 
load(file.path(path, 'Packed_Figure_Reproducibility/input_data_for_EpiTrace/Figure8/matrix/GBM_GSM4131779.testdata.RData')) # this is hg38
annotations <- readRDS(file.path(path,'Packed_Figure_Reproducibility/input_data_for_EpiTrace/Figure8/meta/get_GR_from_EnsDb_hm_v86.rds'))

GBM_GSM4131779$GR -> peakset
initiated_peaks <- Init_Peakset(peakset) 
initiated_mm <- Init_Matrix(cellname = colnames(GBM_GSM4131779$MM),peakname = gsub('_','-',rownames(GBM_GSM4131779$MM)),matrix = GBM_GSM4131779$MM)

epitrace_obj <- EpiTrace_prepare_object(initiated_peaks,initiated_mm,ref_genome = 'hg38',non_standard_clock = F,clock_gr_list = clock_gr_list,qualnum=1)
DefaultAssay(epitrace_obj) <- 'peaks' 
epitrace_obj <- epitrace_obj %>% Signac::RunTFIDF() %>% Signac::FindTopFeatures(min.cutoff=50) %>% Signac::RunSVD() %>% RunUMAP(reduction = 'lsi', dims = 2:31) %>% FindNeighbors(reduction = 'lsi', dims = 2:31,k.param = 21) %>% FindClusters(verbose = FALSE, algorithm = 3,resolution = 0.8)
# epitrace_obj_backup <- epitrace_obj
epitrace_obj2 <- subset(epitrace_obj,seurat_clusters %ni% c('2','8'))

epitrace_obj_age_estimated <- RunEpiTraceAge(epitrace_obj2, normalization_method = 'randomized' )


# Figure 8a
colors_plot <- colorRampPalette(colors=RColorBrewer::brewer.pal(n=9,name = 'Set1'))(length(unique(epitrace_obj_age_estimated$seurat_clusters)))
names(colors_plot) <- unique(epitrace_obj_age_estimated$seurat_clusters)

DimPlot(epitrace_obj_age_estimated,label = T)  -> p1
p1$data %>% dplyr::group_by(ident) %>% dplyr::summarise(umap_1=mean(umap_1),umap_2=mean(umap_2)) -> p1_loc
ggplot(p1$data,aes(x=umap_1,y=umap_2,color=ident)) + geom_point() + theme_classic() + scale_color_manual(values = colors_plot) + theme(legend.position = 'none', text=element_text(size=15)) -> p1

 

fig.path = '/projectnb/ds596/students/preshita/processed_data/all_gbm/GSM4131779_4349/paper_exact_fig'
pdf(file.path(fig.path, "Figure8a_GBM_UMAP.pdf"), width = 8,height = 6)
p1 + geom_label(data=p1_loc,mapping=aes(x=umap_1,y=umap_2,label=ident),fill='white',color='black')
dev.off()

```



```{r}
# Figure 8b to 8e
target_genes <- c('MDM4','EGFR','PDGFRA')
granges(epitrace_obj_age_estimated) -> tempdf_GR
lapply(target_genes,function(x){
  if(x %in% 'PDGFRA'){
    widthid = 10000
  }else{
    widthid = 300
  }
  tempdf_GR[findOverlaps(tempdf_GR,flank(annotations[annotations$gene_name %in% x,],width=widthid,both=T))@from %>% unique,] -> testdf
  testdf$gene <- x
  testdf <- as.data.frame(testdf)
  testdf
}) %>% bind_rows -> testdf 
testdf$target_feature <- paste0(testdf$seqnames,'-',testdf$start,'-',testdf$end)
lapply(target_genes,function(x){
  testdf$target_feature[testdf$gene %in% x]
}) -> featurelist 
names(featurelist) <- target_genes

Seurat::AddModuleScore(epitrace_obj_age_estimated,features = featurelist) -> epitrace_obj_age_estimated
colnames(epitrace_obj_age_estimated@meta.data)[((ncol(epitrace_obj_age_estimated@meta.data)-2):ncol(epitrace_obj_age_estimated@meta.data))] <- target_genes

FeaturePlot(epitrace_obj_age_estimated,features = c('EpiTraceAge_AllClock',target_genes),order = T,ncol = 4,combine = F) -> gene_plot_list 

epitrace_obj_age_estimated@meta.data$seurat_clusters <- factor(epitrace_obj_age_estimated@meta.data$seurat_clusters, levels = rev(c(3,0,5,6,4,7,9,1)))

color_use <- c('coral','chartreuse4','cornflowerblue','red')

lapply(1:length(gene_plot_list),function(idx){
  gene_plot_list[[idx]]$data -> pdata 
  gene_plot_list[[idx]]$labels$title -> ptitle
  if(idx==1){
    ptitle <- 'Clock'
  }
  if(idx==1){
    xlabel <- 'EpiTrace Age'
  }else{
    xlabel <- 'Expression'
  }
  pdata$newvalue <- pdata$value <- pdata[,ncol(pdata)]
  pdata$ident <- factor(pdata$ident,levels=levels(epitrace_obj_age_estimated@meta.data$seurat_clusters),ordered = T)
  ggplot(pdata,aes(x=value,y=ident,fill=ident,group=ident)) + geom_violin(aes(fill=ident),scale='width') + theme_classic() + scale_fill_manual(values = colors_plot) +geom_boxplot(fill='black',width=0.2,outlier.alpha = 0) + theme(legend.position = 'none',axis.title.y = element_blank(),text=element_text(size=15)) + ggtitle(ptitle) + xlab(xlabel)
}) -> final_gene_exp_violin_list  

wrap_plots(final_gene_exp_violin_list,ncol=2) -> p2

pdf(file.path(fig.path,"Figure8b_8e_GBM_expression.pdf"), width = 6,height = 6)
p2
dev.off()

# Figure 8f
epitrace_obj_age_estimated@meta.data -> prepare_data_meta
prepare_data_meta$label <- prepare_data_meta$seurat_clusters
prepare_data_meta %>% dplyr::group_by(label) %>% dplyr::summarise(Normal=sum(EpiTraceAge_MitosisClock<0.3),MDM4_amp=sum(MDM4>-0.1),EGFR=sum(EGFR>0 ),PDGFRA=sum(PDGFRA>0.05),n=n())

prepare_data_meta%>%dplyr::select(EpiTraceAge_MitosisClock,MDM4,PDGFRA,EGFR) %>% dist %>% hclust() %>% cutree(5) -> prepare_data_meta$hdist_class

prepare_data_meta$manual_class <- 'Normal'
prepare_data_meta$manual_class[prepare_data_meta$MDM4 > -0.1] <- 'MDM4+'
prepare_data_meta$manual_class[prepare_data_meta$EGFR > 0 & prepare_data_meta$MDM4 > -0.1] <- 'EGFR+'
prepare_data_meta$manual_class[prepare_data_meta$PDGFRA > 0.1 & prepare_data_meta$MDM4 > -0.1] <- 'PDGFRA+'
prepare_data_meta$manual_class[prepare_data_meta$PDGFRA > 0.1 & prepare_data_meta$EGFR > 0 & prepare_data_meta$MDM4 > -0.1] <- 'EGFR+ PDGFRA+'

prepare_data_meta$manual_class <- factor(prepare_data_meta$manual_class,levels=c('Normal','MDM4+','EGFR+ PDGFRA+','EGFR+','PDGFRA+'))

colors_pie = c('Normal'='white','MDM4+'='gray','EGFR+ PDGFRA+'='orange','EGFR+'='red','PDGFRA+'='black')

p3 = ggplot(prepare_data_meta,aes(x=EpiTraceAge_MitosisClock,y=manual_class)) + geom_violin(aes(fill=manual_class),scale='width') + theme_classic() + geom_boxplot(fill='darkgray',outlier.alpha = 0,width=0.1) + scale_fill_manual(values=colors_pie) + theme(axis.title.y = element_blank(),text=element_text(size=15)) + xlab('MitosisClock')

pdf(file.path(fig.path, "Figure8f_GBM_EpiTrace_Age_in_different_celltypes.pdf"), width = 7,height = 3.5)
p3
dev.off()

# # Figure 8g , tree
# phylotree_res <- RunEpiTracePhylogeny(epitrace_obj_age_estimated,run_reduction = T)
# phylotree_res[[4]][[2]] -> data.tree_clock
# data.tree_clock <- ape::root(data.tree_clock,outgroup=1)
# colors_plot <- colorRampPalette(colors=RColorBrewer::brewer.pal(n=9,name = 'Set1'))(length(unique(epitrace_obj_age_estimated$seurat_clusters)))
# names(colors_plot) <- unique(epitrace_obj_age_estimated$seurat_clusters)
# 
# colors_plot <- colorRampPalette(colors=RColorBrewer::brewer.pal(n=9,name = 'Set1'))(length(unique(prepare_data_meta$seurat_clusters)))
# names(colors_plot) <- unique(prepare_data_meta$seurat_clusters)
# colors_pie = c('Normal'='white','MDM4+'='gray','EGFR+ PDGFRA+'='orange','EGFR+'='red','PDGFRA+'='black')
# 
# p2_tree <- ggtree::ggtree(data.tree_clock,layout='rectangular',ladderize = FALSE)  + geom_tippoint(aes(color = label),size=5) + scale_color_manual(values = colors_plot) + geom_tiplab(color='black',size=4,offset=5) + theme(legend.position='none')
# 
# 
# pdf(file.path(fig.path, "Figure8g_GBM_clock_phylo.pdf"), width = 8,height = 4)
# p2_tree
# dev.off()


# Figure 8g , tree
phylotree_res <- RunEpiTracePhylogeny(epitrace_obj_age_estimated,run_reduction = T)
phylotree_res[[4]][[2]] -> data.tree_clock

# 1. Root the tree (ensure "g1" is the outgroup)
# resolve.root = TRUE is important for a clean root appearance
data.tree_clock <- ape::root(data.tree_clock, outgroup = "g1", resolve.root = TRUE)

# 2. Match colors to the "g" labels
# We need to make sure the names of the color vector match the 'label' column exactly
clusters <- unique(as.character(epitrace_obj_age_estimated$seurat_clusters))
colors_plot <- colorRampPalette(colors=RColorBrewer::brewer.pal(n=9, name = 'Set1'))(length(clusters))
names(colors_plot) <- paste0("g", clusters)

# 3. Plot without restrictive x-limits
library(ggtree)
p2_tree <- ggtree(data.tree_clock, layout='rectangular') + 
  geom_tippoint(aes(color = label), size=5) + 
  geom_tiplab(aes(label = label), color='black', size=4, offset = 0.2) + 
  scale_color_manual(values = colors_plot) + 
  theme(legend.position='right', text=element_text(size=15))

# 4. Automatically adjust the X-axis so labels aren't cut off
p2_tree <- p2_tree + hexpand(0.2, direction = 1) 

p2_tree


pdf(file.path(fig.path, "Figure8g_GBM_clock_phylo.pdf"), width = 8,height = 4)
p2_tree
dev.off()



```


```{r}
library(ggplot2)
library(dplyr)
library(tidyr)

# ── 1. Cluster order matching the tree (top → bottom) ────────────────────────
cluster_order <- c("3", "0", "5", "6", "4", "7", "9", "1")

# ── 2. Build proportion data ──────────────────────────────────────────────────
pie_data <- prepare_data_meta %>%
  dplyr::mutate(seurat_clusters = as.character(seurat_clusters)) %>%
  dplyr::group_by(seurat_clusters, manual_class) %>%
  dplyr::tally() %>%
  dplyr::group_by(seurat_clusters) %>%
  dplyr::mutate(prop = n / sum(n)) %>%
  dplyr::ungroup()

# Fill all cluster × class combinations with 0
all_combos <- expand.grid(
  seurat_clusters  = cluster_order,
  manual_class     = levels(prepare_data_meta$manual_class),
  stringsAsFactors = FALSE
)
pie_data <- dplyr::left_join(all_combos, pie_data,
                              by = c("seurat_clusters", "manual_class"))
pie_data$prop[is.na(pie_data$prop)] <- 0
pie_data$n[is.na(pie_data$n)]       <- 0

# Factor to control facet order
pie_data$seurat_clusters <- factor(pie_data$seurat_clusters,
                                    levels = cluster_order)
pie_data$manual_class    <- factor(pie_data$manual_class,
                                    levels = levels(prepare_data_meta$manual_class))

# ── 3. Plot ───────────────────────────────────────────────────────────────────
p_pies <- ggplot(pie_data %>% dplyr::filter(prop > 0),
                 aes(x = "", y = prop, fill = manual_class)) +

  geom_bar(stat      = "identity",
           width     = 1,
           color     = "black",
           linewidth = 0.4) +

  coord_polar("y", start = 0) +

  facet_wrap(~ seurat_clusters,
             ncol      = 1,          # single vertical column
             strip.position = "left" # cluster label on the left
  ) +

  scale_fill_manual(values = colors_pie, name = "CNV Class") +

  theme_void() +
  theme(
    strip.text       = element_text(size = 14, face = "bold", hjust = 0.5),
    legend.position  = "right",
    legend.title     = element_text(size = 11, face = "bold"),
    legend.text      = element_text(size = 10),
    panel.spacing    = unit(0.1, "lines"),  # gap between pies — reduce to tighten
    plot.margin      = margin(5, 5, 5, 5)
  )

p_pies

# ── 4. Save ───────────────────────────────────────────────────────────────────
pdf(file.path(fig.path, "Figure8g_PieCharts_only.pdf"), width = 3, height = 8)
print(p_pies)
dev.off()
```






Combining all the figures 

```{r}

# Row 1: UMAP (left) | 2×2 violins (right)
row1 <- (p1 + geom_label(data=p1_loc,mapping=aes(x=umap_1,y=umap_2,label=ident),fill='white',color='black') | p2) +
  plot_layout(widths = c(1, 1.1))   # violins slightly wider

# Row 2: MitosisClock violin (left) | Tree (right)
row2 <- (p_f8f | p2_tree) +
  plot_layout(widths = c(1.2, 1))   # violin slightly wider due to y-axis labels

# Stack rows vertically
combined_figure <- (row1 / row2) +
  plot_layout(heights = c(1.2, 1.0)) +  # top row slightly taller
  plot_annotation(
    tag_levels = "a",
    theme = theme(plot.tag = element_text(size = 16, face = "bold"))
  )

# ─────────────────────────────────────────────────────────────────────────────
# Save
# ─────────────────────────────────────────────────────────────────────────────
pdf(file.path(fig.path, "Figure8_combined.pdf"), width = 12, height = 9)
print(combined_figure)
dev.off()

combined_figure
```








# Doing all other data samples


```{r}
# Batch processing for all samples (each in its own output folder)
data_dir <- '/projectnb/ds596/projects/Team 3/data/GSE139136'
main_output_dir <- '/projectnb/ds596/students/preshita/processed_data/all_gbm'
dir.create(main_output_dir, showWarnings = FALSE, recursive = TRUE)

# Load annotation (assume same for all samples)
annotation_path <- '/projectnb/ds596/projects/Team 3/paper_code/Packed_Figure_Reproducibility/input_data_for_EpiTrace/Figure8/meta/get_GR_from_EnsDb_hm_v86.rds'
annotations <- readRDS(annotation_path)

# Helper to get sample prefixes
all_files <- list.files(data_dir, pattern = '_matrix.mtx.gz$', full.names = FALSE)
sample_prefixes <- gsub('_matrix.mtx.gz', '', all_files)

sample_prefixes <- sample_prefixes[-4]
sample_prefixes
```

"GSM4131776_4218" "GSM4131777_4250" "GSM4131778_4275" "GSM4131779_4349"

## Sample 4218 
```{r}
  cat('Processing', sample, '\n')
  matrix_file <- file.path(data_dir, paste0(sample, '_matrix.mtx.gz'))
  peaks_file <- file.path(data_dir, paste0(sample, '_peaks.bed.gz'))
  barcodes_file <- file.path(data_dir, paste0(sample, '_barcodes.tsv.gz'))

  # Create output subfolder for this sample
  sample_output_dir <- file.path(main_output_dir, sample)
  dir.create(sample_output_dir, showWarnings = FALSE, recursive = TRUE)

  # Load data
  mat <- Matrix::readMM(matrix_file)
  peaks <- readr::read_tsv(peaks_file, col_names = FALSE)
  barcodes <- readr::read_tsv(barcodes_file, col_names = FALSE)
  rownames(mat) <- paste0(peaks$X1, '-', peaks$X2, '-', peaks$X3)
  colnames(mat) <- barcodes$X1

  # Build Epitrace object
  peakset <- GenomicRanges::GRanges(seqnames = peaks$X1, ranges = IRanges::IRanges(start = peaks$X2, end = peaks$X3))
  initiated_peaks <- Init_Peakset(peakset)
  initiated_mm <- Init_Matrix(cellname = colnames(mat), peakname = rownames(mat), matrix = mat)
  epitrace_obj <- EpiTrace_prepare_object(initiated_peaks, initiated_mm, ref_genome = 'hg38', non_standard_clock = FALSE, clock_gr_list = clock_gr_list, qualnum = 1)
  DefaultAssay(epitrace_obj) <- 'peaks'
  epitrace_obj <- epitrace_obj %>% Signac::RunTFIDF() %>% Signac::FindTopFeatures(min.cutoff = 50) %>% Signac::RunSVD() %>% RunUMAP(reduction = 'lsi', dims = 2:31) %>% FindNeighbors(reduction = 'lsi', dims = 2:31, k.param = 21) %>% FindClusters(verbose = FALSE, algorithm = 3, resolution = 0.8)


  # Downstream analysis (as before, but save with sample-specific names)
  epitrace_obj_age_estimated <- RunEpiTraceAge(epitrace_obj, normalization_method = 'randomized')

  num_cat = length(unique(epitrace_obj_age_estimated$seurat_clusters))
  # Figure 8a
  colors_plot <- colorRampPalette(colors = RColorBrewer::brewer.pal(n = num_cat, name = 'Set1'))(num_cat)
  names(colors_plot) <- unique(epitrace_obj_age_estimated$seurat_clusters)
  DimPlot(epitrace_obj_age_estimated, label = T)  -> p1
  p1$data %>% dplyr::group_by(ident) %>% dplyr::summarise(umap_1 = mean(umap_1), umap_2 = mean(umap_2)) -> p1_loc
  ggplot(p1$data, aes(x = umap_1, y = umap_2, color = ident)) + geom_point() + theme_classic() + scale_color_manual(values = colors_plot) + theme(legend.position = 'none', axis.text.x = element_text(size = 12), axis.text.y = element_text(size = 15)) -> p1
  pdf(file.path(sample_output_dir, 'Figure8a_UMAP.pdf'), width = 6, height = 6)
  print(p1 + geom_label(data = p1_loc, mapping = aes(x = umap_1, y = umap_2, label = ident), fill = 'white', color = 'black'))
  dev.off()

  # Figure 8b-8e
  target_genes <- c('MDM4', 'EGFR', 'PDGFRA')
  granges(epitrace_obj_age_estimated) -> tempdf_GR
lapply(target_genes, function(x) {
  if (x %in% 'PDGFRA') {
    widthid = 10000
  } else {
    widthid = 300
  }
  
  testdf <- tempdf_GR[findOverlaps(tempdf_GR, flank(annotations[annotations$gene_name %in% x, ], width = widthid, both = T))@from %>% unique, ]
  
  # Check if GRanges has elements
  if (length(testdf) > 0) {
    testdf$gene <- x
    testdf <- as.data.frame(testdf)
    testdf
  } else {
    NULL
  }
}) %>% bind_rows -> testdf
  
testdf$target_feature <- paste0(testdf$seqnames, '-', testdf$start, '-', testdf$end)
  lapply(target_genes, function(x) {
    testdf$target_feature[testdf$gene %in% x]
  }) -> featurelist
  names(featurelist) <- target_genes
  
  
# Filter out empty lists
featurelist <- featurelist[sapply(featurelist, length) > 0]

# Check which genes were removed
removed_genes <- target_genes[!(target_genes %in% names(featurelist))]
if (length(removed_genes) > 0) {
  message("Genes removed (no features found): ", paste(removed_genes, collapse = ", "))
}

# Now run AddModuleScore

  
  Seurat::AddModuleScore(epitrace_obj_age_estimated, features = featurelist) -> epitrace_obj_age_estimated
  colnames(epitrace_obj_age_estimated@meta.data)[((ncol(epitrace_obj_age_estimated@meta.data) - 2):ncol(epitrace_obj_age_estimated@meta.data))] <- target_genes
  FeaturePlot(epitrace_obj_age_estimated, features = c('EpiTraceAge_AllClock', target_genes), order = T, ncol = 4, combine = F) -> gene_plot_list
  epitrace_obj_age_estimated@meta.data$seurat_clusters <- factor(epitrace_obj_age_estimated@meta.data$seurat_clusters, levels = rev(sort(unique(epitrace_obj_age_estimated@meta.data$seurat_clusters))))
  color_use <- c('coral', 'chartreuse4', 'cornflowerblue', 'red')
  lapply(1:length(gene_plot_list), function(idx) {
    gene_plot_list[[idx]]$data -> pdata
    gene_plot_list[[idx]]$labels$title -> ptitle
    if (idx == 1) {
      ptitle <- 'Clock'
      xlabel <- 'EpiTrace Age'
    } else {
      xlabel <- 'Expression'
    }
    pdata$newvalue <- pdata$value <- pdata[, ncol(pdata)]
    pdata$ident <- factor(pdata$ident, levels = levels(epitrace_obj_age_estimated@meta.data$seurat_clusters), ordered = T)
    ggplot(pdata, aes(x = value, y = ident, fill = ident, group = ident)) +
      geom_violin(aes(fill = ident), scale = 'width') +
      theme_classic() +
      scale_fill_manual(values = colors_plot) +
      geom_boxplot(fill = 'black', width = 0.2, outlier.alpha = 0) +
      theme(legend.position = 'none', axis.title.y = element_blank(), text = element_text(size = 15)) +
      ggtitle(ptitle) + xlab(xlabel)
  }) -> final_gene_exp_violin_list
  wrap_plots(final_gene_exp_violin_list, ncol = 2) -> p2
  pdf(file.path(sample_output_dir, 'Figure8b_8e_expression.pdf'), width = 6, height = 6)
  print(p2)
  dev.off()

  
  
```


### panel f and g
For 8f - MDM4 amplification threshold of -0.1 is determined from sample 4349 and applied to all samples.


```{r}
prepare_data_meta <- epitrace_obj_age_estimated@meta.data

# ── 1. Visualize the actual separation ───────────────────────────────────────
# Plot all three features that could separate cluster 7
plot_data <- prepare_data_meta %>%
  dplyr::mutate(
    seurat_clusters = as.character(seurat_clusters),
    group = dplyr::case_when(
      seurat_clusters == "0" ~ "Cluster 0 (Normal)",
      seurat_clusters == "9" ~ "Cluster 9 (Normal)",
      seurat_clusters == "7" ~ "Cluster 7 (Pre-malignant)",
      TRUE                   ~ "Other"
    )
  ) %>%
  dplyr::filter(group != "Other")

# MDM4 vs Clock scatter — this will show whether clock separates them
p_scatter <- ggplot(plot_data,
                    aes(x = MDM4,
                        y = EpiTraceAge_MitosisClock,
                        color = group)) +
  geom_point(alpha = 0.4, size = 1) +
  scale_color_manual(values = c(
    "Cluster 0 (Normal)"      = "steelblue",
    "Cluster 9 (Normal)"      = "cornflowerblue",
    "Cluster 7 (Pre-malignant)" = "orange"
  )) +
  theme_classic() +
  theme(legend.position = "top", legend.title = element_blank()) +
  labs(title = "MDM4 score vs MitosisClock\n(only clusters 0, 7, 9)",
       x = "MDM4 score", y = "MitosisClock")

# Clock density for three clusters
p_clock_density <- ggplot(plot_data,
                           aes(x = EpiTraceAge_MitosisClock, fill = group)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c(
    "Cluster 0 (Normal)"        = "steelblue",
    "Cluster 9 (Normal)"        = "cornflowerblue",
    "Cluster 7 (Pre-malignant)" = "orange"
  )) +
  theme_classic() +
  theme(legend.position = "top", legend.title = element_blank()) +
  labs(title = "MitosisClock density\n(the only real separator)",
       x = "MitosisClock")

# MDM4 density for three clusters
p_mdm4_density <- ggplot(plot_data,
                          aes(x = MDM4, fill = group)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c(
    "Cluster 0 (Normal)"        = "steelblue",
    "Cluster 9 (Normal)"        = "cornflowerblue",
    "Cluster 7 (Pre-malignant)" = "orange"
  )) +
  theme_classic() +
  theme(legend.position = "top", legend.title = element_blank()) +
  labs(title = "MDM4 score density\n(nearly identical → cannot separate)",
       x = "MDM4 score")

(p_mdm4_density | p_clock_density | p_scatter)
# → You will see MDM4 densities overlap completely
# → Clock is the ONLY feature that separates cluster 7
```



```{r}
# ── 2. Find the clock threshold that separates Normal from cluster 7 ──────────
clock_normal   <- prepare_data_meta %>%
  dplyr::mutate(seurat_clusters = as.character(seurat_clusters)) %>%
  dplyr::filter(seurat_clusters %in% c("0", "9")) %>%
  dplyr::pull(EpiTraceAge_MitosisClock)

clock_cluster7 <- prepare_data_meta %>%
  dplyr::mutate(seurat_clusters = as.character(seurat_clusters)) %>%
  dplyr::filter(seurat_clusters == "7") %>%
  dplyr::pull(EpiTraceAge_MitosisClock)

# Midpoint between 95th pct of Normal and 5th pct of cluster 7
clock_thresh <- mean(c(
  quantile(clock_normal,   0.95, na.rm = TRUE),
  quantile(clock_cluster7, 0.05, na.rm = TRUE)
))

cat(sprintf("Clock threshold to separate Normal from cluster 7: %.4f\n", clock_thresh))

# Also use 90th pct of Normal cells (clock + gene) for gene thresholds
normal_cells <- prepare_data_meta %>%
  dplyr::mutate(seurat_clusters = as.character(seurat_clusters)) %>%
  dplyr::filter(seurat_clusters %in% c("0", "9"))

MDM4_thresh   <- -0.1  # threshold from sample 4349
EGFR_thresh   <- quantile(normal_cells$EGFR,   0.90, na.rm = TRUE)
PDGFRA_thresh <- quantile(normal_cells$PDGFRA, 0.90, na.rm = TRUE)

cat(sprintf("MDM4   threshold: %.4f\n", MDM4_thresh))
cat(sprintf("EGFR   threshold: %.4f\n", EGFR_thresh))
cat(sprintf("PDGFRA threshold: %.4f\n", PDGFRA_thresh))
```


```{r}
# ── 3. Apply classification: Normal requires LOW clock AND LOW gene scores ─────
prepare_data_meta <- prepare_data_meta %>%
  dplyr::mutate(seurat_clusters = as.character(seurat_clusters))

# Start everyone as Normal
prepare_data_meta$manual_class <- "Normal"

# Apply gene-score-based classes
prepare_data_meta$manual_class[
  prepare_data_meta$MDM4 > MDM4_thresh
] <- "MDM4+"

prepare_data_meta$manual_class[
  prepare_data_meta$EGFR > EGFR_thresh &
  prepare_data_meta$MDM4 > MDM4_thresh
] <- "EGFR+"

prepare_data_meta$manual_class[
  prepare_data_meta$PDGFRA > PDGFRA_thresh &
  prepare_data_meta$MDM4   > MDM4_thresh
] <- "PDGFRA+"

prepare_data_meta$manual_class[
  prepare_data_meta$PDGFRA > PDGFRA_thresh &
  prepare_data_meta$EGFR   > EGFR_thresh   &
  prepare_data_meta$MDM4   > MDM4_thresh
] <- "EGFR+ PDGFRA+"

# KEY: Cells with "Normal" gene scores but HIGH clock → reclassify as MDM4+
# This captures cluster 7 without a forced override
prepare_data_meta$manual_class[
  prepare_data_meta$manual_class == "Normal" &       # passed gene score filter
  prepare_data_meta$EpiTraceAge_MitosisClock > clock_thresh  # but clock is elevated
] <- "MDM4+"                                         # → reclassify as MDM4+

prepare_data_meta$manual_class <- factor(
  prepare_data_meta$manual_class,
  levels = c("Normal", "MDM4+", "EGFR+ PDGFRA+", "EGFR+", "PDGFRA+")
)
```


```{r}
# ── 5. Plot p3 ────────────────────────────────────────────────────────────────
colors_pie <- c(
  "Normal"        = "white",
  "MDM4+"         = "gray",
  "EGFR+ PDGFRA+" = "orange",
  "EGFR+"         = "red",
  "PDGFRA+"       = "black"
)

p3 <- ggplot(prepare_data_meta,
             aes(x = EpiTraceAge_MitosisClock, y = manual_class)) +
  geom_violin(aes(fill = manual_class), scale = "width") +
  geom_boxplot(fill = "darkgray", outlier.alpha = 0, width = 0.1) +
  scale_fill_manual(values = colors_pie) +
  theme_classic() +
  theme(axis.title.y = element_blank(),
        text         = element_text(size = 15)) +
  xlab("MitosisClock")

p3

pdf(file.path(sample_output_dir, "Figure8f_EpiTrace_Age_in_celltypes.pdf"),
    width = 7, height = 3.5)
print(p3)
dev.off()


  saveRDS(epitrace_obj_age_estimated, file.path(sample_output_dir, 'epitrace_obj_age_estimated.rds'))
  
```


Making phylotree separately
```{r}
# Figure 8g , tree
phylotree_res <- RunEpiTracePhylogeny(epitrace_obj_age_estimated,run_reduction = T)
phylotree_res[[4]][[2]] -> data.tree_clock

# 1. Root the tree (ensure "g1" is the outgroup)
# resolve.root = TRUE is important for a clean root appearance
data.tree_clock <- ape::root(data.tree_clock, outgroup = "g0", resolve.root = TRUE)

# 2. Match colors to the "g" labels
# We need to make sure the names of the color vector match the 'label' column exactly
clusters <- unique(as.character(epitrace_obj_age_estimated$seurat_clusters))
colors_plot <- colorRampPalette(colors=RColorBrewer::brewer.pal(n=9, name = 'Set1'))(length(clusters))
names(colors_plot) <- paste0("g", clusters)

# 3. Plot without restrictive x-limits
library(ggtree)
p2_tree <- ggtree(data.tree_clock, layout='rectangular') + 
  geom_tippoint(aes(color = label), size=5) + 
  geom_tiplab(aes(label = label), color='black', size=4, offset = 0.2) + 
  scale_color_manual(values = colors_plot) + 
  theme(legend.position='right', text=element_text(size=15))

# 4. Automatically adjust the X-axis so labels aren't cut off
p2_tree <- p2_tree + hexpand(0.2, direction = 1) 

p2_tree


pdf(file.path(sample_output_dir, "Figure8g_GBM_clock_phylo.pdf"), width = 8,height = 4)
p2_tree
dev.off()
```



Making pie charts separately
```{r}
library(ggplot2)
library(dplyr)
library(tidyr)

# ── 1. Cluster order matching the tree (top → bottom) ────────────────────────
cluster_order <- c("8", "1", "5", "6", "3", "4", "2", "7", "9", "0")

# ── 2. Build proportion data ──────────────────────────────────────────────────
pie_data <- prepare_data_meta %>%
  dplyr::mutate(seurat_clusters = as.character(seurat_clusters)) %>%
  dplyr::group_by(seurat_clusters, manual_class) %>%
  dplyr::tally() %>%
  dplyr::group_by(seurat_clusters) %>%
  dplyr::mutate(prop = n / sum(n)) %>%
  dplyr::ungroup()

# Fill all cluster × class combinations with 0
all_combos <- expand.grid(
  seurat_clusters  = cluster_order,
  manual_class     = levels(prepare_data_meta$manual_class),
  stringsAsFactors = FALSE
)
pie_data <- dplyr::left_join(all_combos, pie_data,
                              by = c("seurat_clusters", "manual_class"))
pie_data$prop[is.na(pie_data$prop)] <- 0
pie_data$n[is.na(pie_data$n)]       <- 0

# Factor to control facet order
pie_data$seurat_clusters <- factor(pie_data$seurat_clusters,
                                    levels = cluster_order)
pie_data$manual_class    <- factor(pie_data$manual_class,
                                    levels = levels(prepare_data_meta$manual_class))

# ── 3. Plot ───────────────────────────────────────────────────────────────────
p_pies <- ggplot(pie_data %>% dplyr::filter(prop > 0),
                 aes(x = "", y = prop, fill = manual_class)) +

  geom_bar(stat      = "identity",
           width     = 1,
           color     = "black",
           linewidth = 0.4) +

  coord_polar("y", start = 0) +

  facet_wrap(~ seurat_clusters,
             ncol      = 1,          # single vertical column
             strip.position = "left" # cluster label on the left
  ) +

  scale_fill_manual(values = colors_pie, name = "CNV Class") +

  theme_void() +
  theme(
    strip.text       = element_text(size = 14, face = "bold", hjust = 0.5),
    legend.position  = "right",
    legend.title     = element_text(size = 11, face = "bold"),
    legend.text      = element_text(size = 10),
    panel.spacing    = unit(0.1, "lines"),  # gap between pies — reduce to tighten
    plot.margin      = margin(5, 5, 5, 5)
  )

p_pies

# ── 4. Save ───────────────────────────────────────────────────────────────────
pdf(file.path(sample_output_dir, "Figure8g_PieCharts_only.pdf"), width = 3, height = 8)
print(p_pies)
dev.off()
```






Combining all the figures 
```{r}

# Row 1: UMAP (left) | 2×2 violins (right)
row1 <- (p1 + geom_label(data=p1_loc,mapping=aes(x=umap_1,y=umap_2,label=ident),fill='white',color='black') | p2) +
  plot_layout(widths = c(1, 1.1))   # violins slightly wider

# Row 2: MitosisClock violin (left) | Tree (right)
row2 <- (p3 | p2_tree) +
  plot_layout(widths = c(1.2, 1))   # violin slightly wider due to y-axis labels

# Stack rows vertically
combined_figure <- (row1 / row2) +
  plot_layout(heights = c(1.2, 1.0)) +  # top row slightly taller
  plot_annotation(
    tag_levels = "a",
    theme = theme(plot.tag = element_text(size = 16, face = "bold"))
  )

# ─────────────────────────────────────────────────────────────────────────────
# Save
# ─────────────────────────────────────────────────────────────────────────────
pdf(file.path(sample_output_dir, "Figure8_combined.pdf"), width = 12, height = 9)
print(combined_figure)
dev.off()

combined_figure
```




## Sample 4250
```{r}
sample <- 'GSM4131777_4250'
cat('Processing', sample, '\n')
matrix_file <- file.path(data_dir, paste0(sample, '_matrix.mtx.gz'))
peaks_file <- file.path(data_dir, paste0(sample, '_peaks.bed.gz'))
barcodes_file <- file.path(data_dir, paste0(sample, '_barcodes.tsv.gz'))

# Create output subfolder for this sample
sample_output_dir <- file.path(main_output_dir, sample)
dir.create(sample_output_dir, showWarnings = FALSE, recursive = TRUE)

# Load data
mat <- Matrix::readMM(matrix_file)
peaks <- readr::read_tsv(peaks_file, col_names = FALSE)
barcodes <- readr::read_tsv(barcodes_file, col_names = FALSE)
rownames(mat) <- paste0(peaks$X1, '-', peaks$X2, '-', peaks$X3)
colnames(mat) <- barcodes$X1

# Build Epitrace object
peakset <- GenomicRanges::GRanges(seqnames = peaks$X1, ranges = IRanges::IRanges(start = peaks$X2, end = peaks$X3))
initiated_peaks <- Init_Peakset(peakset)
initiated_mm <- Init_Matrix(cellname = colnames(mat), peakname = rownames(mat), matrix = mat)
epitrace_obj <- EpiTrace_prepare_object(initiated_peaks, initiated_mm, ref_genome = 'hg38', non_standard_clock = FALSE, clock_gr_list = clock_gr_list, qualnum = 1)
DefaultAssay(epitrace_obj) <- 'peaks'
epitrace_obj <- epitrace_obj %>% Signac::RunTFIDF() %>% Signac::FindTopFeatures(min.cutoff = 50) %>% Signac::RunSVD() %>% RunUMAP(reduction = 'lsi', dims = 2:31) %>% FindNeighbors(reduction = 'lsi', dims = 2:31, k.param = 21) %>% FindClusters(verbose = FALSE, algorithm = 3, resolution = 0.8)

# Downstream analysis
epitrace_obj_age_estimated <- RunEpiTraceAge(epitrace_obj, normalization_method = 'randomized')

num_cat = length(unique(epitrace_obj_age_estimated$seurat_clusters))
# Figure 8a
colors_plot <- colorRampPalette(colors = RColorBrewer::brewer.pal(n = num_cat, name = 'Set1'))(num_cat)
names(colors_plot) <- unique(epitrace_obj_age_estimated$seurat_clusters)
DimPlot(epitrace_obj_age_estimated, label = T)  -> p1
p1$data %>% dplyr::group_by(ident) %>% dplyr::summarise(umap_1 = mean(umap_1), umap_2 = mean(umap_2)) -> p1_loc
ggplot(p1$data, aes(x = umap_1, y = umap_2, color = ident)) + geom_point() + theme_classic() + scale_color_manual(values = colors_plot) + theme(legend.position = 'none', axis.text.x = element_text(size = 12), axis.text.y = element_text(size = 15)) -> p1
pdf(file.path(sample_output_dir, 'Figure8a_UMAP.pdf'), width = 6, height = 6)
print(p1 + geom_label(data = p1_loc, mapping = aes(x = umap_1, y = umap_2, label = ident), fill = 'white', color = 'black'))
dev.off()

# Figure 8b-8e
target_genes <- c('MDM4', 'EGFR', 'PDGFRA')
granges(epitrace_obj_age_estimated) -> tempdf_GR
lapply(target_genes, function(x) {
  if (x %in% 'PDGFRA') {
    widthid = 10000
  } else {
    widthid = 300
  }
  testdf <- tempdf_GR[findOverlaps(tempdf_GR, flank(annotations[annotations$gene_name %in% x, ], width = widthid, both = T))@from %>% unique, ]
  if (length(testdf) > 0) {
    testdf$gene <- x
    testdf <- as.data.frame(testdf)
    testdf
  } else {
    NULL
  }
}) %>% bind_rows -> testdf

testdf$target_feature <- paste0(testdf$seqnames, '-', testdf$start, '-', testdf$end)
lapply(target_genes, function(x) {
  testdf$target_feature[testdf$gene %in% x]
}) -> featurelist
names(featurelist) <- target_genes

featurelist <- featurelist[sapply(featurelist, length) > 0]
removed_genes <- target_genes[!(target_genes %in% names(featurelist))]
if (length(removed_genes) > 0) {
  message("Genes removed (no features found): ", paste(removed_genes, collapse = ", "))
}

Seurat::AddModuleScore(epitrace_obj_age_estimated, features = featurelist) -> epitrace_obj_age_estimated
colnames(epitrace_obj_age_estimated@meta.data)[((ncol(epitrace_obj_age_estimated@meta.data) - 2):ncol(epitrace_obj_age_estimated@meta.data))] <- target_genes
FeaturePlot(epitrace_obj_age_estimated, features = c('EpiTraceAge_AllClock', target_genes), order = T, ncol = 4, combine = F) -> gene_plot_list
epitrace_obj_age_estimated@meta.data$seurat_clusters <- factor(epitrace_obj_age_estimated@meta.data$seurat_clusters, levels = rev(sort(unique(epitrace_obj_age_estimated@meta.data$seurat_clusters))))
color_use <- c('coral', 'chartreuse4', 'cornflowerblue', 'red')
lapply(1:length(gene_plot_list), function(idx) {
  gene_plot_list[[idx]]$data -> pdata
  gene_plot_list[[idx]]$labels$title -> ptitle
  if (idx == 1) {
    ptitle <- 'Clock'
    xlabel <- 'EpiTrace Age'
  } else {
    xlabel <- 'Expression'
  }
  pdata$newvalue <- pdata$value <- pdata[, ncol(pdata)]
  pdata$ident <- factor(pdata$ident, levels = levels(epitrace_obj_age_estimated@meta.data$seurat_clusters), ordered = T)
  ggplot(pdata, aes(x = value, y = ident, fill = ident, group = ident)) +
    geom_violin(aes(fill = ident), scale = 'width') +
    theme_classic() +
    scale_fill_manual(values = colors_plot) +
    geom_boxplot(fill = 'black', width = 0.2, outlier.alpha = 0) +
    theme(legend.position = 'none', axis.title.y = element_blank(), text = element_text(size = 15)) +
    ggtitle(ptitle) + xlab(xlabel)
}) -> final_gene_exp_violin_list
wrap_plots(final_gene_exp_violin_list, ncol = 2) -> p2
pdf(file.path(sample_output_dir, 'Figure8b_8e_expression.pdf'), width = 6, height = 6)
print(p2)
dev.off()
```


### panel f and g for 4250
```{r}
prepare_data_meta <- epitrace_obj_age_estimated@meta.data

# MDM4 threshold based on sample 4349
MDM4_thresh <- -0.1

# Compute clock and other gene thresholds from normal clusters
clock_normal <- prepare_data_meta %>%
  dplyr::mutate(seurat_clusters = as.character(seurat_clusters)) %>%
  dplyr::filter(seurat_clusters %in% c("0", "9")) %>%
  dplyr::pull(EpiTraceAge_MitosisClock)

clock_cluster7 <- prepare_data_meta %>%
  dplyr::mutate(seurat_clusters = as.character(seurat_clusters)) %>%
  dplyr::filter(seurat_clusters == "7") %>%
  dplyr::pull(EpiTraceAge_MitosisClock)

clock_thresh <- mean(c(
  quantile(clock_normal,   0.95, na.rm = TRUE),
  quantile(clock_cluster7, 0.05, na.rm = TRUE)
))

cat(sprintf("Clock threshold to separate Normal from cluster 7: %.4f\n", clock_thresh))

normal_cells <- prepare_data_meta %>%
  dplyr::mutate(seurat_clusters = as.character(seurat_clusters)) %>%
  dplyr::filter(seurat_clusters %in% c("0", "9"))

EGFR_thresh   <- quantile(normal_cells$EGFR,   0.90, na.rm = TRUE)
PDGFRA_thresh <- quantile(normal_cells$PDGFRA, 0.90, na.rm = TRUE)

cat(sprintf("MDM4   threshold: %.4f\n", MDM4_thresh))
cat(sprintf("EGFR   threshold: %.4f\n", EGFR_thresh))
cat(sprintf("PDGFRA threshold: %.4f\n", PDGFRA_thresh))
```


```{r}
prepare_data_meta <- prepare_data_meta %>%
  dplyr::mutate(seurat_clusters = as.character(seurat_clusters))

prepare_data_meta$manual_class <- "Normal"

prepare_data_meta$manual_class[
  prepare_data_meta$MDM4 > MDM4_thresh
] <- "MDM4+"

prepare_data_meta$manual_class[
  prepare_data_meta$EGFR > EGFR_thresh &
  prepare_data_meta$MDM4 > MDM4_thresh
] <- "EGFR+"

prepare_data_meta$manual_class[
  prepare_data_meta$PDGFRA > PDGFRA_thresh &
  prepare_data_meta$MDM4   > MDM4_thresh
] <- "PDGFRA+"

prepare_data_meta$manual_class[
  prepare_data_meta$PDGFRA > PDGFRA_thresh &
  prepare_data_meta$EGFR   > EGFR_thresh   &
  prepare_data_meta$MDM4   > MDM4_thresh
] <- "EGFR+ PDGFRA+"

prepare_data_meta$manual_class[
  prepare_data_meta$manual_class == "Normal" &
  prepare_data_meta$EpiTraceAge_MitosisClock > clock_thresh
] <- "MDM4+"

prepare_data_meta$manual_class <- factor(
  prepare_data_meta$manual_class,
  levels = c("Normal", "MDM4+", "EGFR+ PDGFRA+", "EGFR+", "PDGFRA+")
)
```


```{r}
colors_pie <- c(
  "Normal"        = "white",
  "MDM4+"         = "gray",
  "EGFR+ PDGFRA+" = "orange",
  "EGFR+"         = "red",
  "PDGFRA+"       = "black"
)

p3 <- ggplot(prepare_data_meta,
             aes(x = EpiTraceAge_MitosisClock, y = manual_class)) +
  geom_violin(aes(fill = manual_class), scale = "width") +
  geom_boxplot(fill = "darkgray", outlier.alpha = 0, width = 0.1) +
  scale_fill_manual(values = colors_pie) +
  theme_classic() +
  theme(axis.title.y = element_blank(),
        text         = element_text(size = 15)) +
  xlab("MitosisClock")

p3

pdf(file.path(sample_output_dir, "Figure8f_EpiTrace_Age_in_celltypes.pdf"),
    width = 7, height = 3.5)
print(p3)
dev.off()

saveRDS(epitrace_obj_age_estimated, file.path(sample_output_dir, 'epitrace_obj_age_estimated.rds'))
```

Making phylotree separately
```{r}
phylotree_res <- RunEpiTracePhylogeny(epitrace_obj_age_estimated, run_reduction = T)
phylotree_res[[4]][[2]] -> data.tree_clock

data.tree_clock <- ape::root(data.tree_clock, outgroup = "g0", resolve.root = TRUE)

clusters <- unique(as.character(epitrace_obj_age_estimated$seurat_clusters))
colors_plot <- colorRampPalette(colors=RColorBrewer::brewer.pal(n=9, name = 'Set1'))(length(clusters))
names(colors_plot) <- paste0("g", clusters)

library(ggtree)
p2_tree <- ggtree(data.tree_clock, layout='rectangular') +
  geom_tippoint(aes(color = label), size=5) +
  geom_tiplab(aes(label = label), color='black', size=4, offset = 0.2) +
  scale_color_manual(values = colors_plot) +
  theme(legend.position='right', text=element_text(size=15))

p2_tree <- p2_tree + hexpand(0.2, direction = 1)

p2_tree

pdf(file.path(sample_output_dir, "Figure8g_GBM_clock_phylo.pdf"), width = 8, height = 4)
p2_tree
dev.off()
```


Making pie charts separately
```{r}
library(ggplot2)
library(dplyr)
library(tidyr)

cluster_order <- rev(sort(unique(as.character(prepare_data_meta$seurat_clusters))))

pie_data <- prepare_data_meta %>%
  dplyr::mutate(seurat_clusters = as.character(seurat_clusters)) %>%
  dplyr::group_by(seurat_clusters, manual_class) %>%
  dplyr::tally() %>%
  dplyr::group_by(seurat_clusters) %>%
  dplyr::mutate(prop = n / sum(n)) %>%
  dplyr::ungroup()

all_combos <- expand.grid(
  seurat_clusters  = cluster_order,
  manual_class     = levels(prepare_data_meta$manual_class),
  stringsAsFactors = FALSE
)
pie_data <- dplyr::left_join(all_combos, pie_data,
                              by = c("seurat_clusters", "manual_class"))
pie_data$prop[is.na(pie_data$prop)] <- 0
pie_data$n[is.na(pie_data$n)]       <- 0

pie_data$seurat_clusters <- factor(pie_data$seurat_clusters, levels = cluster_order)
pie_data$manual_class    <- factor(pie_data$manual_class, levels = levels(prepare_data_meta$manual_class))

p_pies <- ggplot(pie_data %>% dplyr::filter(prop > 0),
                 aes(x = "", y = prop, fill = manual_class)) +
  geom_bar(stat = "identity", width = 1, color = "black", linewidth = 0.4) +
  coord_polar("y", start = 0) +
  facet_wrap(~ seurat_clusters, ncol = 1, strip.position = "left") +
  scale_fill_manual(values = colors_pie, name = "CNV Class") +
  theme_void() +
  theme(
    strip.text       = element_text(size = 14, face = "bold", hjust = 0.5),
    legend.position  = "right",
    legend.title     = element_text(size = 11, face = "bold"),
    legend.text      = element_text(size = 10),
    panel.spacing    = unit(0.1, "lines"),
    plot.margin      = margin(5, 5, 5, 5)
  )

p_pies

pdf(file.path(sample_output_dir, "Figure8g_PieCharts_only.pdf"), width = 3, height = 8)
print(p_pies)
dev.off()
```


Combining all the figures
```{r}
row1 <- (p1 + geom_label(data=p1_loc, mapping=aes(x=umap_1, y=umap_2, label=ident), fill='white', color='black') | p2) +
  plot_layout(widths = c(1, 1.1))

row2 <- (p3 | p2_tree) +
  plot_layout(widths = c(1.2, 1))

combined_figure <- (row1 / row2) +
  plot_layout(heights = c(1.2, 1.0)) +
  plot_annotation(
    tag_levels = "a",
    theme = theme(plot.tag = element_text(size = 16, face = "bold"))
  )

pdf(file.path(sample_output_dir, "Figure8_combined.pdf"), width = 12, height = 9)
print(combined_figure)
dev.off()

combined_figure
```




## Sample 4275
```{r}
sample <- 'GSM4131778_4275'
cat('Processing', sample, '\n')
matrix_file <- file.path(data_dir, paste0(sample, '_matrix.mtx.gz'))
peaks_file <- file.path(data_dir, paste0(sample, '_peaks.bed.gz'))
barcodes_file <- file.path(data_dir, paste0(sample, '_barcodes.tsv.gz'))

# Create output subfolder for this sample
sample_output_dir <- file.path(main_output_dir, sample)
dir.create(sample_output_dir, showWarnings = FALSE, recursive = TRUE)

# Load data
mat <- Matrix::readMM(matrix_file)
peaks <- readr::read_tsv(peaks_file, col_names = FALSE)
barcodes <- readr::read_tsv(barcodes_file, col_names = FALSE)
rownames(mat) <- paste0(peaks$X1, '-', peaks$X2, '-', peaks$X3)
colnames(mat) <- barcodes$X1

# Build Epitrace object
peakset <- GenomicRanges::GRanges(seqnames = peaks$X1, ranges = IRanges::IRanges(start = peaks$X2, end = peaks$X3))
initiated_peaks <- Init_Peakset(peakset)
initiated_mm <- Init_Matrix(cellname = colnames(mat), peakname = rownames(mat), matrix = mat)
epitrace_obj <- EpiTrace_prepare_object(initiated_peaks, initiated_mm, ref_genome = 'hg38', non_standard_clock = FALSE, clock_gr_list = clock_gr_list, qualnum = 1)
DefaultAssay(epitrace_obj) <- 'peaks'
epitrace_obj <- epitrace_obj %>% Signac::RunTFIDF() %>% Signac::FindTopFeatures(min.cutoff = 50) %>% Signac::RunSVD() %>% RunUMAP(reduction = 'lsi', dims = 2:31) %>% FindNeighbors(reduction = 'lsi', dims = 2:31, k.param = 21) %>% FindClusters(verbose = FALSE, algorithm = 3, resolution = 0.8)

# Downstream analysis
epitrace_obj_age_estimated <- RunEpiTraceAge(epitrace_obj, normalization_method = 'randomized')

num_cat = length(unique(epitrace_obj_age_estimated$seurat_clusters))
# Figure 8a
colors_plot <- colorRampPalette(colors = RColorBrewer::brewer.pal(n = num_cat, name = 'Set1'))(num_cat)
names(colors_plot) <- unique(epitrace_obj_age_estimated$seurat_clusters)
DimPlot(epitrace_obj_age_estimated, label = T)  -> p1
p1$data %>% dplyr::group_by(ident) %>% dplyr::summarise(umap_1 = mean(umap_1), umap_2 = mean(umap_2)) -> p1_loc
ggplot(p1$data, aes(x = umap_1, y = umap_2, color = ident)) + geom_point() + theme_classic() + scale_color_manual(values = colors_plot) + theme(legend.position = 'none', axis.text.x = element_text(size = 12), axis.text.y = element_text(size = 15)) -> p1
pdf(file.path(sample_output_dir, 'Figure8a_UMAP.pdf'), width = 6, height = 6)
print(p1 + geom_label(data = p1_loc, mapping = aes(x = umap_1, y = umap_2, label = ident), fill = 'white', color = 'black'))
dev.off()

# Figure 8b-8e
target_genes <- c('MDM4', 'EGFR', 'PDGFRA')
granges(epitrace_obj_age_estimated) -> tempdf_GR
lapply(target_genes, function(x) {
  if (x %in% 'PDGFRA') {
    widthid = 10000
  } else {
    widthid = 300
  }
  testdf <- tempdf_GR[findOverlaps(tempdf_GR, flank(annotations[annotations$gene_name %in% x, ], width = widthid, both = T))@from %>% unique, ]
  if (length(testdf) > 0) {
    testdf$gene <- x
    testdf <- as.data.frame(testdf)
    testdf
  } else {
    NULL
  }
}) %>% bind_rows -> testdf

testdf$target_feature <- paste0(testdf$seqnames, '-', testdf$start, '-', testdf$end)
lapply(target_genes, function(x) {
  testdf$target_feature[testdf$gene %in% x]
}) -> featurelist
names(featurelist) <- target_genes

featurelist <- featurelist[sapply(featurelist, length) > 0]
removed_genes <- target_genes[!(target_genes %in% names(featurelist))]
if (length(removed_genes) > 0) {
  message("Genes removed (no features found): ", paste(removed_genes, collapse = ", "))
}

Seurat::AddModuleScore(epitrace_obj_age_estimated, features = featurelist) -> epitrace_obj_age_estimated
colnames(epitrace_obj_age_estimated@meta.data)[((ncol(epitrace_obj_age_estimated@meta.data) - 2):ncol(epitrace_obj_age_estimated@meta.data))] <- target_genes
FeaturePlot(epitrace_obj_age_estimated, features = c('EpiTraceAge_AllClock', target_genes), order = T, ncol = 4, combine = F) -> gene_plot_list
epitrace_obj_age_estimated@meta.data$seurat_clusters <- factor(epitrace_obj_age_estimated@meta.data$seurat_clusters, levels = rev(sort(unique(epitrace_obj_age_estimated@meta.data$seurat_clusters))))
color_use <- c('coral', 'chartreuse4', 'cornflowerblue', 'red')
lapply(1:length(gene_plot_list), function(idx) {
  gene_plot_list[[idx]]$data -> pdata
  gene_plot_list[[idx]]$labels$title -> ptitle
  if (idx == 1) {
    ptitle <- 'Clock'
    xlabel <- 'EpiTrace Age'
  } else {
    xlabel <- 'Expression'
  }
  pdata$newvalue <- pdata$value <- pdata[, ncol(pdata)]
  pdata$ident <- factor(pdata$ident, levels = levels(epitrace_obj_age_estimated@meta.data$seurat_clusters), ordered = T)
  ggplot(pdata, aes(x = value, y = ident, fill = ident, group = ident)) +
    geom_violin(aes(fill = ident), scale = 'width') +
    theme_classic() +
    scale_fill_manual(values = colors_plot) +
    geom_boxplot(fill = 'black', width = 0.2, outlier.alpha = 0) +
    theme(legend.position = 'none', axis.title.y = element_blank(), text = element_text(size = 15)) +
    ggtitle(ptitle) + xlab(xlabel)
}) -> final_gene_exp_violin_list
wrap_plots(final_gene_exp_violin_list, ncol = 2) -> p2
pdf(file.path(sample_output_dir, 'Figure8b_8e_expression.pdf'), width = 6, height = 6)
print(p2)
dev.off()
```


### panel f and g for 4275
```{r}
prepare_data_meta <- epitrace_obj_age_estimated@meta.data

# MDM4 threshold based on sample 4349
MDM4_thresh <- -0.1

# Compute clock and other gene thresholds from normal clusters
clock_normal <- prepare_data_meta %>%
  dplyr::mutate(seurat_clusters = as.character(seurat_clusters)) %>%
  dplyr::filter(seurat_clusters %in% c("0", "9")) %>%
  dplyr::pull(EpiTraceAge_MitosisClock)

clock_cluster7 <- prepare_data_meta %>%
  dplyr::mutate(seurat_clusters = as.character(seurat_clusters)) %>%
  dplyr::filter(seurat_clusters == "7") %>%
  dplyr::pull(EpiTraceAge_MitosisClock)

clock_thresh <- mean(c(
  quantile(clock_normal,   0.95, na.rm = TRUE),
  quantile(clock_cluster7, 0.05, na.rm = TRUE)
))

cat(sprintf("Clock threshold to separate Normal from cluster 7: %.4f\n", clock_thresh))

normal_cells <- prepare_data_meta %>%
  dplyr::mutate(seurat_clusters = as.character(seurat_clusters)) %>%
  dplyr::filter(seurat_clusters %in% c("0", "9"))

EGFR_thresh   <- quantile(normal_cells$EGFR,   0.90, na.rm = TRUE)
PDGFRA_thresh <- quantile(normal_cells$PDGFRA, 0.90, na.rm = TRUE)

cat(sprintf("MDM4   threshold: %.4f\n", MDM4_thresh))
cat(sprintf("EGFR   threshold: %.4f\n", EGFR_thresh))
cat(sprintf("PDGFRA threshold: %.4f\n", PDGFRA_thresh))
```


```{r}
prepare_data_meta <- prepare_data_meta %>%
  dplyr::mutate(seurat_clusters = as.character(seurat_clusters))

prepare_data_meta$manual_class <- "Normal"

prepare_data_meta$manual_class[
  prepare_data_meta$MDM4 > MDM4_thresh
] <- "MDM4+"

prepare_data_meta$manual_class[
  prepare_data_meta$EGFR > EGFR_thresh &
  prepare_data_meta$MDM4 > MDM4_thresh
] <- "EGFR+"

prepare_data_meta$manual_class[
  prepare_data_meta$PDGFRA > PDGFRA_thresh &
  prepare_data_meta$MDM4   > MDM4_thresh
] <- "PDGFRA+"

prepare_data_meta$manual_class[
  prepare_data_meta$PDGFRA > PDGFRA_thresh &
  prepare_data_meta$EGFR   > EGFR_thresh   &
  prepare_data_meta$MDM4   > MDM4_thresh
] <- "EGFR+ PDGFRA+"

prepare_data_meta$manual_class[
  prepare_data_meta$manual_class == "Normal" &
  prepare_data_meta$EpiTraceAge_MitosisClock > clock_thresh
] <- "MDM4+"

prepare_data_meta$manual_class <- factor(
  prepare_data_meta$manual_class,
  levels = c("Normal", "MDM4+", "EGFR+ PDGFRA+", "EGFR+", "PDGFRA+")
)
```


```{r}
colors_pie <- c(
  "Normal"        = "white",
  "MDM4+"         = "gray",
  "EGFR+ PDGFRA+" = "orange",
  "EGFR+"         = "red",
  "PDGFRA+"       = "black"
)

p3 <- ggplot(prepare_data_meta,
             aes(x = EpiTraceAge_MitosisClock, y = manual_class)) +
  geom_violin(aes(fill = manual_class), scale = "width") +
  geom_boxplot(fill = "darkgray", outlier.alpha = 0, width = 0.1) +
  scale_fill_manual(values = colors_pie) +
  theme_classic() +
  theme(axis.title.y = element_blank(),
        text         = element_text(size = 15)) +
  xlab("MitosisClock")

p3

pdf(file.path(sample_output_dir, "Figure8f_EpiTrace_Age_in_celltypes.pdf"),
    width = 7, height = 3.5)
print(p3)
dev.off()

saveRDS(epitrace_obj_age_estimated, file.path(sample_output_dir, 'epitrace_obj_age_estimated.rds'))
```

Making phylotree separately
```{r}
phylotree_res <- RunEpiTracePhylogeny(epitrace_obj_age_estimated, run_reduction = T)
phylotree_res[[4]][[2]] -> data.tree_clock

data.tree_clock <- ape::root(data.tree_clock, outgroup = "g0", resolve.root = TRUE)

clusters <- unique(as.character(epitrace_obj_age_estimated$seurat_clusters))
colors_plot <- colorRampPalette(colors=RColorBrewer::brewer.pal(n=9, name = 'Set1'))(length(clusters))
names(colors_plot) <- paste0("g", clusters)

library(ggtree)
p2_tree <- ggtree(data.tree_clock, layout='rectangular') +
  geom_tippoint(aes(color = label), size=5) +
  geom_tiplab(aes(label = label), color='black', size=4, offset = 0.2) +
  scale_color_manual(values = colors_plot) +
  theme(legend.position='right', text=element_text(size=15))

p2_tree <- p2_tree + hexpand(0.2, direction = 1)

p2_tree

pdf(file.path(sample_output_dir, "Figure8g_GBM_clock_phylo.pdf"), width = 8, height = 4)
p2_tree
dev.off()
```


Making pie charts separately
```{r}
library(ggplot2)
library(dplyr)
library(tidyr)

cluster_order <- rev(sort(unique(as.character(prepare_data_meta$seurat_clusters))))

pie_data <- prepare_data_meta %>%
  dplyr::mutate(seurat_clusters = as.character(seurat_clusters)) %>%
  dplyr::group_by(seurat_clusters, manual_class) %>%
  dplyr::tally() %>%
  dplyr::group_by(seurat_clusters) %>%
  dplyr::mutate(prop = n / sum(n)) %>%
  dplyr::ungroup()

all_combos <- expand.grid(
  seurat_clusters  = cluster_order,
  manual_class     = levels(prepare_data_meta$manual_class),
  stringsAsFactors = FALSE
)
pie_data <- dplyr::left_join(all_combos, pie_data,
                              by = c("seurat_clusters", "manual_class"))
pie_data$prop[is.na(pie_data$prop)] <- 0
pie_data$n[is.na(pie_data$n)]       <- 0

pie_data$seurat_clusters <- factor(pie_data$seurat_clusters, levels = cluster_order)
pie_data$manual_class    <- factor(pie_data$manual_class, levels = levels(prepare_data_meta$manual_class))

p_pies <- ggplot(pie_data %>% dplyr::filter(prop > 0),
                 aes(x = "", y = prop, fill = manual_class)) +
  geom_bar(stat = "identity", width = 1, color = "black", linewidth = 0.4) +
  coord_polar("y", start = 0) +
  facet_wrap(~ seurat_clusters, ncol = 1, strip.position = "left") +
  scale_fill_manual(values = colors_pie, name = "CNV Class") +
  theme_void() +
  theme(
    strip.text       = element_text(size = 14, face = "bold", hjust = 0.5),
    legend.position  = "right",
    legend.title     = element_text(size = 11, face = "bold"),
    legend.text      = element_text(size = 10),
    panel.spacing    = unit(0.1, "lines"),
    plot.margin      = margin(5, 5, 5, 5)
  )

p_pies

pdf(file.path(sample_output_dir, "Figure8g_PieCharts_only.pdf"), width = 3, height = 8)
print(p_pies)
dev.off()
```


Combining all the figures
```{r}
row1 <- (p1 + geom_label(data=p1_loc, mapping=aes(x=umap_1, y=umap_2, label=ident), fill='white', color='black') | p2) +
  plot_layout(widths = c(1, 1.1))

row2 <- (p3 | p2_tree) +
  plot_layout(widths = c(1.2, 1))

combined_figure <- (row1 / row2) +
  plot_layout(heights = c(1.2, 1.0)) +
  plot_annotation(
    tag_levels = "a",
    theme = theme(plot.tag = element_text(size = 16, face = "bold"))
  )

pdf(file.path(sample_output_dir, "Figure8_combined.pdf"), width = 12, height = 9)
print(combined_figure)
dev.off()

combined_figure
```



